---
title: 'Soil microbial biodiversity: Bacteria abundant taxa'
author: "Mercedes Roman"
date: "`r Sys.Date()`"
output: word_document
---

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}

library(rgdal)
library(gdalUtils)
library(raster)
library(sp)
library(sf)
library(dplyr)
#library(tidyverse)
library(tidyr)
library(ggmap)
library(ggplot2) 
library(viridis) # color palettes
library(scales)
library(rasterVis)
library(lattice)
library(gridExtra)
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(mapview) # for interactive maps
library(shiny)   # for web applications
library(foreach)
library(doParallel)
library(geosphere)
library(gplots)
library(colorspace)
library(readr)
library(vegan)
library(readr)
library(phyloseq)
library(speedyseq)
library(ecoCopula)
library(metagMisc)

### Load the data
InputDir <- "R:/PRJ-AusSoilMicrobes/Output/Bacteria/"
load(paste0(InputDir,"2_Rare_Abun_Bacteria.RData" ))
rm(rare.ps)

### Load Phylogenetic tree calculated in another script
load(paste0(InputDir,"Abun_Phyloseq_and_Tree_16S.RData"))

### And 
load("C:/Users/mrom8073/OneDrive - The University of Sydney (Staff)/SoilMicrobialBiodiversity/Presentations/Bac.Abun.Rdata.RData")

### Also the total counts
InputDir <- "R:/PRJ-AusSoilMicrobes/Output/Bacteria/"
load(paste0(InputDir,"1_AbsoluteAbun.RData"))

InputDir <- "R:/PRJ-AusSoilMicrobes/Output/Bacteria/"
OutDir <- "R:/PRJ-AusSoilMicrobes/Output/Bacteria/"

```

## Soil bacteria - Abundant taxa

We transformed the abundance data into relative abundance and calculated the average relative abundance of each OTU across all sites. The abundant taxa were defined as those from the top 1%, after ranking the OTUs by relative abundance, that occur in at least 10 % sites of sites or those OTUs that occurred in 50% of sites.The dominant taxa accounted mostly between 10-30 % of the OTUs present in the samples.Theese criteria resulted in 2063 taxa accross 1373 samples.

```{r}
abun.ps
```
We will test three methods for analyzing the biodiversity: non-metric multidimensional scaling (NMDS), Copula ordination, and Umap.


### NMDS

First I perform a NMDS using Bray-Curtis dissimilarity metric on abundant taxa.

```{r}

set.seed(1812)
Bacteria.Abun.NMDS = ordinate(abun.ps, "NMDS", "bray",                
                           ### Bray-Curtis dissimilarity
                           k=3,                                       
                           ### 3 dimensions
                           try=20, trymax=2000,                       
                           ### Min 20 random starts, max 2000
                           maxit=5000,                                
                           ### Try 5000 iterations at each start
                           smin = 0.05,                               
                           ### I would accept a stress of 0.1, the structure of the community is well represented so I set the threshold in 0.05.
                           trace = FALSE, plot=FALSE)
Bacteria.Abun.NMDS
save(Bacteria.Abun.NMDS,file=paste0(OutDir,"Bacteria.Abun.NMDS.RData"))

```

```{r NMDS_plot, echo=FALSE, message=FALSE, warning=FALSE}

### Let's see the geographical pattern
library(gridExtra)
theme_set(theme_bw())
p1 = plot_ordination(abun.ps, Bacteria.Abun.NMDS, type="taxa", color="Phylum", title="taxa") + scale_color_viridis(discrete = TRUE, option="D", direction = -1)
p2 = plot_ordination(abun.ps, Bacteria.Abun.NMDS, type="samples", color="Organic.Carbon....") + scale_color_viridis(discrete = FALSE, option="A", direction = -1)
p3 = plot_ordination(abun.ps, Bacteria.Abun.NMDS, type="samples", color="Ph.Solid.H2o") + scale_color_viridis(discrete = FALSE,option="A", direction = -1)
p4 = plot_ordination(abun.ps, axes = c(1,3), Bacteria.Abun.NMDS, type="samples", color="Ph.Solid.H2o") + scale_color_viridis(discrete = FALSE,option="A", direction = -1)
p1
p2
p3
p4

```

The stress of the NMDS is 0.11, suggesting that the structure of the data has been captured relatively well.


```{r ausmap, message=FALSE, warning=FALSE, include=FALSE}

### Plot NMDS results (coordinates)
AusMap <- get_stamenmap(bbox = c(left=110, bottom=-45, right=157, top=-8),
                        maptype="toner-lite", zoom=5,
                        source="stamen", crop=TRUE)

```

There are some spatial patterns, which would be interesting to know if they respond to environmental facotrs, latitude gradient, or dataset of origin (BASE or GA).

```{r NMDS_map_points, echo=FALSE, message=FALSE, warning=FALSE}

ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=NMDS1), data = p2$data) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=NMDS2), data = p2$data) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
#grid.arrange(m1, m2, ncol = 2, nrow = 1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=NMDS3), data = p4$data) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)

```


#### NMDS with weighted unifrac distance

Instead of running the NMDS with weighted unifrac distance directly in physoleq, I first calculate the weighted unifrac distance with phyloseq and the NMDS with the vegan package so I can set my preferred parameters. This NMDS analysis does find a convergent solution with a stress of 0.09, which indicates that the structure of the data has been captured well.

```{r wunifrac nmds vegan, echo=TRUE, message=FALSE, warning=FALSE}

### Plot the tree
plot(fitGTR$tree, cex=0.1)


cl <- makeCluster(6)
registerDoParallel(cl)
set.seed(2665)
uf_BacAbun <- UniFrac(Bac_tree, weighted=TRUE, normalized=TRUE, parallel=TRUE, fast=TRUE)
stopCluster(cl)

class(uf_BacAbun)
dim(uf_BacAbun)

set.seed(5436)
nmds.wuf <- vegan::metaMDS(comm = uf_BacAbun, k = 3, 
                           try = 50, trymax = 2000,
                           trace = FALSE, 
                           plot = FALSE, 
                           maxit=2000,
                           smin = 0.05)

nmds.wuf

ordiplot(nmds.wuf)

```



```{r echo=FALSE}

### Extract scores
nmds.scores <- nmds.wuf$points
context.abun.bac <- sample_data(Bac_tree)
nmds.wuf.context <- cbind(nmds.scores,context.abun.bac)

### Plot the scores on a map
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=MDS1), data = nmds.wuf.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
### eliminate the outlier in MDS2
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=MDS2), data = nmds.wuf.context[nmds.wuf.context$MDS2<0.4,]) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=MDS3), data = nmds.wuf.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)


save(uf_BacAbun,nmds.wuf,nmds.wuf.context, file=paste0(OutDir,"Bacteria.nmds.wuf.RData"))

```

## Copulas models

We start by bringing the total counts 

```{r otu_abun, echo=TRUE}

ps_srf.r

abunA.ps <- prune_taxa(taxa_names(ps_srf.r)%in% taxa_names(Bac_tree),ps_srf.r)
abunA.ps <- prune_samples(sample_sums(abunA.ps)>0, abunA.ps)
abunA.ps

#Prepare the data in the format for the copulas models
ta <- abunA.ps %>% otu_table %>% as.matrix() %>% t() %>% as.data.frame()
ta[1:10,1:10]

context.data <- data.frame(sample_data(abunA.ps))
context.data$Dataset <- ifelse(is.na(context.data$Ph.Solid.H2o), "GA", "BASE")
rownames(context.data) <- 1:nrow(context.data)
rownames(ta) <- 1:nrow(ta)

```

```{r negative binomial, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

context.data$seqs<- log(rowSums(ta))

sdm_Nbinomial <- 
  stackedsdm(ta, formula_X = ~1 + offset(seqs),
                 context.data,
                 family="negative.binomial", ncores = 7 )

sdm_Nbinomial_2 <- 
  stackedsdm(ta, formula_X = ~1 + Depth_Int + Dataset + offset(seqs),
                 context.data,
                 family="negative.binomial", ncores = 7 )
gc()

```

```{r copula_ordination1, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

## plot residuals of models
#(sdm_Nbinomial)

# Fit copula ordination 
abun_lv=cord(sdm_Nbinomial)

```


```{r plot_ccord, eval=FALSE, include=FALSE}
#Plot
data.origin <- ifelse(context.data$Dataset=="GA", "darkcyan", "orange") 
plot(abun_lv, site.col=data.origin)
```
```{r include=FALSE}
gc()
```

We can check the optimal number of latent factors, up to a number that we consider may be useful to map (maximum 5). We check the BIC of the ordination for 2 to 5 latent factors.

```{r optimal_lv, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(purrr)

#max_lv<- ncol(ra)-1 

lvs<- c(2,3,4,5)

cords<-
  map(lvs,
      function(lv){
        cord(sdm_Nbinomial, nlv=lv)
})

best_fit = map_dbl(cords, ~ .x$BIC) %>% which.min()

map_dbl(cords, ~ .x$BIC)
print(best_fit)


```

#### Visualize the latent factors

```{r ltf_map_points, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

abun_lv=cord(sdm_Nbinomial)
ltnV <- as.data.frame(abun_lv$scores)
ltnV <- cbind(ltnV, context.data)

m1 <- ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Factor1), data = ltnV) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1);m1
m2 <- ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Factor2), data = ltnV) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1);m2

```



## UMAP

```{r umap_wunifrac, echo=TRUE, message=FALSE, warning=FALSE}

library(umap)

### Tune parameters
custom.config = umap.defaults
custom.config$n_components=3
custom.config$random_state = 1984
bacteria.umap = umap(as.matrix(uf_BacAbun),config=custom.config, input="dist")

### Plot
umap.context <- sample_data(Bac_tree)
umap.context$Dataset <- ifelse(is.na(umap.context$Ph.Solid.H2o), "GA", "BASE")
umap.context$Dataset <- as.factor(umap.context$Dataset)

### Attach the umap scores
umap.context$Umap1 <- bacteria.umap$layout[,1]
umap.context$Umap2 <- bacteria.umap$layout[,2]
umap.context$Umap3 <- bacteria.umap$layout[,3]
#with(nmds.wuf.context, levels(Dataset))

data.origin <- ifelse(umap.context$Dataset=="GA", "darkcyan", "orange") 

plot(x=umap.context$Umap1, y=umap.context$Umap2 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP2")
plot(x=umap.context$Umap1, y=umap.context$Umap3 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP3")

save(bacteria.umap,umap.context, file=paste0(OutDir,"Bacteria.umap.wuf.RData"))

```


```{r umap_wunifrac_map}

ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap1), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap2), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap3), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)


```


```{r umap_bray, echo=TRUE, message=FALSE, warning=FALSE}

### I calculate first Bray distances

#Prepare the relative abundance in the right format
ra <- Bac_tree %>% otu_table %>% as.matrix() %>% t() %>% as.data.frame()
ra[1:10,1:10]

bray.dist <- vegan::vegdist(ra, "bray")

library(umap)
### Tune parameters
custom.config = umap.defaults
custom.config$n_components=3
custom.config$random_state = 1984

### Perform UMAP
bacteria.umap.bray = umap(as.matrix(bray.dist),config=custom.config, input="dist")

### Plot
umap.context <- sample_data(Bac_tree)
umap.context$Dataset <- ifelse(is.na(umap.context$Ph.Solid.H2o), "GA", "BASE")
umap.context$Dataset <- as.factor(umap.context$Dataset)

### Attach the umap scores
umap.context$Umap1 <- bacteria.umap.bray$layout[,1]
umap.context$Umap2 <- bacteria.umap.bray$layout[,2]
umap.context$Umap3 <- bacteria.umap.bray$layout[,3]
#with(nmds.wuf.context, levels(Dataset))

data.origin <- ifelse(umap.context$Dataset=="GA", "darkcyan", "orange") 

plot(x=umap.context$Umap1, y=umap.context$Umap2 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP2",
     main="UMAP visualization with Bray distance - Relative abundance Bacteria")
plot(x=umap.context$Umap1, y=umap.context$Umap3 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP3",
     main="UMAP visualization with Bray distance - Relative abundance Bacteria")

save(bacteria.umap.bray,umap.context, file=paste0(OutDir,"Bacteria.umap.bray.RData"))

```




```{r umap_bray_map}

ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap1), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap2), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap3), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)

```




```{r  umap_ra_Eculidean, echo=TRUE, message=FALSE, warning=FALSE}

### Tune parameters
custom.config = umap.defaults
custom.config$n_components=3
custom.config$random_state = 1984

bacteria.umap2 = umap(ra, config=custom.config)

### Plot
umap.context <- sample_data(Bac_tree)
umap.context$Dataset <- ifelse(is.na(umap.context$Ph.Solid.H2o), "GA", "BASE")
umap.context$Dataset <- as.factor(umap.context$Dataset)

### Attach the umap scores
umap.context$Umap1 <- bacteria.umap2$layout[,1]
umap.context$Umap2 <- bacteria.umap2$layout[,2]
umap.context$Umap3 <- bacteria.umap2$layout[,3]

data.origin <- ifelse(umap.context$Dataset=="GA", "darkcyan", "orange") 

plot(x=umap.context$Umap1, y=umap.context$Umap2 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP2",
     main="UMAP visualization with Euclidean distance - Relative abundance bacteria")
plot(x=umap.context$Umap1, y=umap.context$Umap3 , col=data.origin, pch=19, xlab="UMAP 1", ylab="UMAP3",
     main="UMAP visualization with Euclidean distance - Relative abundance bacteria")

save(bacteria.umap2,umap.context, file=paste0(OutDir,"Bacteria.umap.eucl.RData"))

```




```{r umap_ra_Euclidean_map}

ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap1), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap2), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)
ggmap(AusMap) + geom_point(aes(y = Latitude, x =Longitude,color=Umap3), data = umap.context) +
  scale_color_viridis(discrete = FALSE, option="A", direction = -1)

```


#### End of the script for now
